Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Rest of said page...

Given such a substitution \(\gamma\), there are two ways to use it. The first is
standard capture-avoiding substitution, which we denote \([\gamma]e\) for
expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for CFGs, as is
standard. The second is to use the substitution to generate a set of
let-bindings for a context \(\Gamma\), written
\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
CFGs. In Section \ref{sec:semantics}, we will show that this is semantically
equivalent to regular capture-avoiding substitution, due to the fact that all
terms in a substitution are central \textit{and} that only affine computations
can be used in a term substituting for an affine variable, and only relevant
computationbs can be used in a term substituting for a relevant variable.

@@

To do this, we begin by introducing the notion of a \textit{symmetric
premonoidal subcategory}, as follows:
\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
\end{definition} 
Note that the set of symmetric premonoidal subcategories, monoidal
subcategories, and central subcategories of a category \(\mc{C}\) are all closed
under intersections and therefore, in particular, form a complete
meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where
\(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
largest central subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\); the
existence of a top-element means that in these cases we may induce the structure
of a complete lattice in the usual manner.

We can now define a \textit{substructural premonoidal category}, which is simply
a premonoidal category equipped with a distinguished set of relevant and affine
morphisms
\begin{definition}[Substructural (Pre)monoidal Category]
  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
  \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\ms{r}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
    \end{itemize}
  \end{itemize}
  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
A simple example of a substructural Cartesian category is the category of
relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad
with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular,
we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^{\ms{a}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\ms{r}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and
\(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to
prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and
that they satisfy the axioms of a commutative comonoid, as desired.

>>>>

Given such a substitution \(\gamma\), there are two ways to use it. The first is
standard capture-avoiding substitution, which we denote \([\gamma]e\) for
expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for CFGs, as is
standard. The second is to use the substitution to generate a set of
let-bindings for a context \(\Gamma\), written
\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
\ref{fig:ssa-subst-ops}. 
%
% Note that this operation cannot be easily defined for CFGs. 
%
% In Section \ref{sec:semantics}, we will show that this is semantically
% equivalent to regular capture-avoiding substitution, due to the fact that all
% terms in a substitution are central \textit{and} that only affine computations
% can be used in a term substituting for an affine variable, and only relevant
% computationbs can be used in a term substituting for a relevant variable.

@@

To do this, we begin by introducing the notion of a \textit{symmetric
premonoidal subcategory}, as follows:
\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
\end{definition}

Note that the set of symmetric premonoidal subcategories, monoidal
subcategories, and central subcategories of a category \(\mc{C}\) are all closed
under intersections.
%
% and therefore, in particular, form a complete meet-semilattice with shared
% bottom element \(\ms{Skeleton}(\mc{C})\), where \(\ms{Skeleton}(C)\) is the
% smallest possible symmetric premonodial subcategory. The largest symmetric
% premonoidal subcategory is of course \(\mc{C}\), while the largest central
% subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\).
%
% ; the existence of a top-element means that in these cases we may induce the
% structure of a complete lattice in the usual manner.
%
We can now define a \textit{substructural premonoidal category}, which is simply
a premonoidal category equipped with a distinguished set of relevant and affine
morphisms
\begin{definition}[Substructural (Pre)monoidal Category]
  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
  \mc{R})\) where:
    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
    closed under tensor products, such that \(\mb{1} \in \mc{A}\) and
      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
    closed under tensor products, such that \(\mb{1} \in \mc{R}\) and
      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\ms{r}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
A simple example of a substructural Cartesian category is the category of
relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad
with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular,
we can define:
  \begin{aligned}
    \ms{Rel}^{\ms{a}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\ms{r}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and
\(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to
prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and
that they satisfy the axioms of a commutative comonoid, as desired.
